\prefacesection{Abstract}

The convergence of optical character recognition (OCR), natural language processing, and generative artificial intelligence has created unprecedented opportunities for developing intelligent content transformation systems. This thesis presents the design, implementation, and evaluation of an integrated system that combines custom-trained Tesseract OCR models, locally deployed large language models, and state-of-the-art AI-driven image generation to create a comprehensive text-to-image transformation pipeline.

The system architecture employs a hybrid approach that balances local processing for privacy-sensitive operations with cloud-based services for computationally intensive image generation tasks. The OCR component utilizes custom-trained Tesseract models with advanced preprocessing techniques including adaptive thresholding, contrast enhancement, and noise reduction, achieving character-level accuracy improvements of 8.3\% over baseline implementations for degraded images. The natural language processing module leverages the GPT-OSS-20B model deployed locally through the Ollama framework, providing sophisticated text understanding and enhancement capabilities while maintaining complete data privacy through local processing.

The image generation subsystem integrates with FLUX 1.1 Pro Ultra, a 12-billion parameter diffusion model that incorporates flow matching techniques and hybrid multimodal transformers. The system achieves exceptional prompt adherence scores of 9.1/10 with average generation times of 24.3 seconds, demonstrating superior performance compared to alternative approaches. The modular architecture ensures seamless coordination between components while maintaining system responsiveness and comprehensive error handling.

Comprehensive evaluation across multiple performance dimensions validates the effectiveness of the integrated approach. The system demonstrates robust performance across diverse content types, maintaining high accuracy for both printed and handwritten text recognition, effective prompt optimization that enhances semantic richness by 91\%, and high-quality image synthesis with strong visual coherence. The research contributes significant advances in custom OCR model training methodologies, local large language model deployment strategies, and multimodal AI system integration patterns.

The practical implementation as a native macOS application demonstrates the feasibility of deploying sophisticated AI capabilities on consumer hardware while maintaining acceptable performance characteristics. This work establishes important precedents for privacy-preserving multimodal AI applications and provides a comprehensive framework for future research in intelligent content transformation systems.